Practical Hilbert space approximate Bayesian Gaussian processes for probabilistic programming

### 2022-03-12 Paper/Manuscript_Stats&Comp/manuscript_STCO.tex
last edits

Equations in Section~\ref{sec_num_equations} to update $m$ and $c$ imply:
% 
\begin{itemize}
	\item $c$ must be big enough for a given $\ell$, and
	\item $m$ must be big enough for given $\ell$ and $c$. 
\end{itemize}
%
If larger than minimal $c$ and $m$ (for a given $\ell$) are used in the initial HSGP model, it is likely that the results are already sufficiently accurate. As $\ell$ is initially unknown, we recommend using this diagnostic in an iterative procedure by starting with $c$ and $m$ based on some initial guess about $\ell$, and if the estimated $\hat{\ell}$ is below the diagnostic threshold, select new $c$ and $m$ using $\hat{\ell}$. This can be repeated until
%
\begin{itemize}
\item the estimated $\hat{\ell}$ is larger than the diagnostic threshold given $c$ and $m$, and
\item the predictive accuracy measures, for example, root mean square error (RMSE), coefficient of determination ($R^2$), or expected log predicitve density (ELPD) don't improve.
\end{itemize}

As commented above, the estimated $\hat{\ell}$ being larger than diagnostic threshold doesn't guarantee that the approximate is sufficiently accurate, and thus we recommend to look at the predicitve accuracy measures, too.

Apart from providing a powerful diagnostic tool in determining if the approxiamtion is sufficiently accurate, the equations in the previous Section~\ref{sec_num_equations} also provide the optimal values for $m$ (the minimum $m$ required for an accurate approximation) and $c$ (the minimum $c$ that allows for the minimum $m$) that van be used to minimize the computational cost in repeated computations (e.g., in cross-validation and simulation based calibration)
%If the initial lengthscale is not far from the optimal, and larger than minimal $c$ and $m$ are used in the initial HSGP model, it is likely that the initial results are already sufficiently accurate. 
%The iterative procedure, briefly commented above and detailed in next Section~\ref{sec_user_guide}, works as safety approach and also allows finding a smaller sufficient $m$ which can be useful if the user wants to optimize the computational speed.
%Using the optimum values for $m$ and $c$ that allow for minimum computational requirements can be essential for repeated fitting (e.g., in cross-validation or simulation based calibration), uni-dimensional functions with very large data sets, and
This even more useful in multi-dimensional cases ($2 \leq D \leq 4$), where knowing the smallest useful value of $m$ for each dimension has even bigger effect on the total computational cost.

@@

The HSGP model is computationally superior in $1$D and $2$D even for highly wiggly functions, except when the number of data points so small ($n \lesssim 300$) that exact GPs are already reasonably fast themselves. However, the computation time of the HSGP model increases rapidly with the number of input dimensions ($D$) since the number of multivariate basis functions $m^*=m_1\times \cdots \times m_D$ in the approximation increases exponentially with $D$ (see eq.~(\ref{eq_m_multi})). Yet, the HSGP method can still be computationally faster than the exact GP for larger datasets due the latter's cubic scaling in $n$.

In our experiments of multivariate problems (see Section~\ref{sec_computation_nD}), the computation time for the HSGP model was faster than for the exact GP for most of the non-linear $2D$ functions or moderate-to-large sized $3$D datasets ($n>1000$), even for highly wiggly functions with large $m^* \approx 10000$ (see Figure~\ref{fig_time_nD_log}).

For small sized datasets ($n<1000$), HSGPs are likely to be slower than exact GPs already for $D=3$ and $D=4$, with a relatively large number of multivariate basis functions ($m^* > 3000$) in $3$D, and with a relatively low number of multivariate basis functions ($m^* \gtrsim 3000$) in $4$D (see Figure~\ref{fig_time_nD_log}).

@@

As it has been shown in case study III (Section~\ref{sec_caseIII}), the proposed diagnostic tool can be very useful for multivariate problems as it allows one to reduce $m^*$ to the minimum sufficient value, reducing computational time drastically, and still getting an accurate approximation.% Otherwise, the user might be struggling in these multivariate cases.

@@

In our experiments, we observed roughly similar or even worse scaling behavior for the thin plate splines in comparison to HSGPs, with serious challenges with computation time when doing inference with spline models for $D=3$ and more than 10 basis functions, or even for $D=2$ and more than 40 basis functions (see Figures~B.3-right and C.3 in the online supplement).

@@

The joint posterior parameter distributions are estimated by sampling using the dynamic HMC algorithm implemented in Stan \citep{StanTeam:2021}. $\mathrm{Normal}(0,1)$, $\mathrm{Normal}(0,3)$ and $\mathrm{Gamma}(1.2,0.2)$ prior distributions has been used for the observation noise $\sigma$, covariance function marginal variance $\alpha$, and length-scale $\ell$, respectively. We use the same prior distributions to fit the exact GP model.

@@

The Stan model code for the exact GP, the approximate GP and the spline models of this case study can be found online at {\small \url{ https://github.com/gabriuma/basis_functions_approach_to_GP/tree/master/Paper/Case-study_1D-Simulated-data}}\,.

@@

The joint posterior parameter distributions are estimated by sampling using the dynamic HMC algorithm implemented in Stan \citep{StanTeam:2021}. $\mathrm{Normal}(0,1)$, $\mathrm{Normal}(0,10)$ and $\mathrm{Normal}(0,2)$ prior distributions has been used for the observation noise $\sigma$, covariance function marginal variances $\bm{\alpha}=\{\alpha_1 \alpha_2, \alpha_3\}$, and length-scales $\bm{\ell}=\{\ell, \ell,\ell\}$, respectively. A $\mathrm{Normal}(0,0.1)$ prior distribution has been used for the standard deviation $\tau$ of the Student's $t$ distribution with 1 degree of freedom used to model $f_4$ (i.e., the special days effects).

@@

The Stan model code for the approximate GP model of this case study can be found online at {\small \url{https://github.com/gabriuma/basis_functions_approach_to_GP/tree/master/Paper/Case-study_Birthday-data}}\,.

@@

In this case study, we apply the diagnostic tool to fit and diagnose two different data sets, one data set simulated from a bivariate (D=$2$) function and another data set simulated from a $3$-dimensinal (D=$3$) function. Furthermore, we present results or the computational time required to fit the HSGP model in $2$D, $3$D and $4$D input spaces, with different sized data sets and as a function of the number of multivariate basis functions $m^*=m_1\times \cdots \times m_D$ used in the approximation.

The $2$D and $3$D synthetic datasets consist of either $n=200$ and $n=1000$ ($i=1,\dots,n$) individual draws from GP priors with input values $\bm{x}_i \in \{[-1,1],\dots,[-1,1]\} \subset {\rm I\!R}^D$, in a D-dimensional input space. We use a GP prior with a squared exponential covariance function with  marginal variance $\alpha=1$ and length-scales $\ell_1=0.10$, $\ell_2=0.3$, and $\ell_3=0.4$, where $\ell_i$ is the length scale for the $i$'th dimension. Independent Gaussian noise with standard deviation $\sigma=0.2$ was added to the GP draws to form the final noisy sets of observations.

@@

\begin{table}
\centering
\setlength{\tabcolsep}{3.6pt}
\begin{tabular}{cccccccccc}
\arrayrulecolor{gray}\hline \\[-3mm]
 iter. & Dim. & $\ell$ & $c$ & $m$ & $\hat{\ell}$ & $\hat{\ell} + 0.01 \geq \ell$ & RMSE & RMSE* \\ 
\arrayrulecolor{lightgray}\hline \\[-2mm]
 \multirow{2}{*}{ 1 } & $1$ & 0.50 & 1.6 & 6 & 0.19 & FALSE & \multirow{2}{*}{ 0.75 } & \multirow{2}{*}{ 0.72 }\\ 
 & $2$ & 1.00 & 3.2 & 6 & 0.25 & FALSE &  &  \\
\arrayrulecolor{lightgray}\hline \\[-2mm]
 \multirow{2}{*}{ 2 } & $1$ & 0.19 & 1.2 & 12 & 0.11 & FALSE  & \multirow{2}{*}{ 0.18 } & \multirow{2}{*}{ 0.12 }\\
 & $2$ & 0.25 & 1.2 & 9 & 0.25 & TRUE &  &  \\
\arrayrulecolor{lightgray}\hline \\[-2mm]
 \multirow{2}{*}{ 3 } & $1$ & 0.11 & 1.2 & 20 & 0.11 & TRUE  & \multirow{2}{*}{ 0.11 } & \multirow{2}{*}{ 0.03 }\\
 & $2$ & 0.19 & 1.2 & 11 & 0.27 & TRUE &  &  \\
\arrayrulecolor{lightgray}\hline \\[-2mm]
 \multirow{2}{*}{ 4 } & $1$ & 0.10 & 1.2 & 22 & 0.11 & TRUE  & \multirow{2}{*}{ 0.10 } & \multirow{2}{*}{ 0.02 }\\
 & $2$ & 0.19 & 1.2 & 11 & 0.27 & TRUE  &  & \\[1mm] 
\arrayrulecolor{lightgray}\hline
 \multicolumn{8}{l}{True lengthscales: $\ell_{1_{GP}}=0.10$, \, $\ell_{2_{GP}}=0.29$ }
\end{tabular}
\caption{Iterative steps in HSGP model fitting on the $2$D data. The diagnostic procedure updating the parameters minimum $\ell$, $c$, $m$, estimated $\hat{\ell}$, and diagnostic $\hat{\ell} + 0.01 \geq \ell$ is applied on the two input dimensions. The full diagnostic process is applied until two consecutive \textit{trues} in the diagnostic are achieved for each dimension. RMSE and RMSE* refer to the root mean squared error of the HSGP computed against the true function and the exact GP, respectively.}
  \label{tab_caseIII_2D}
\end{table}

\begin{table}
\centering
\setlength{\tabcolsep}{3.6pt}
\begin{tabular}{cccccccccc}
\arrayrulecolor{gray}\hline \\[-3mm]
 iter. & Dim. & $\ell$ & $c$ & $m$ & $\hat{\ell}$ & $\hat{\ell} + 0.01 \geq \ell$ & RMSE & RMSE* \\ 
\arrayrulecolor{lightgray}\hline \\[-2mm]
 \multirow{3}{*}{ 1 } & $1$ & 0.50 & 1.6 & 6 & 0.14 & FALSE & \multirow{3}{*}{ 0.63 } & \multirow{3}{*}{ 0.58 }\\ 
 & $2$ & 0.50 & 1.6 & 6 & 0.27 & FALSE &  &  \\
 & $3$ & 1.00 & 3.2 & 6 & 0.23 & FALSE &  &  \\
\arrayrulecolor{lightgray}\hline \\[-2mm]
 \multirow{3}{*}{ 2 } & $1$ & 0.13 & 1.2 & 16 & 0.11 & FALSE  & \multirow{3}{*}{ 0.13 } & \multirow{3}{*}{ 0.07 }\\
 & $2$ & 0.27 & 1.2 & 8 & 0.29 & TRUE &  &  \\
 & $3$ & 0.23 & 1.2 & 19 & 0.40 & TRUE &  &  \\
\arrayrulecolor{lightgray}\hline \\[-2mm]
 \multirow{3}{*}{ 3 } & $1$ & 0.11 & 1.2 & 19 & 0.12 & TRUE  & \multirow{3}{*}{ 0.11 } & \multirow{3}{*}{ 0.04 }\\
 & $2$ & 0.21 & 1.2 & 10 & 0.29 & TRUE &  &  \\
 & $3$ & 0.19 & 1.3 & 12 & 0.43 & TRUE &  &  \\
 \arrayrulecolor{lightgray}\hline \\[-2mm]
 \multirow{3}{*}{ 4 } & $1$ & 0.10 & 1.2 & 21 & 0.12 & TRUE  & \multirow{3}{*}{ 0.11 } & \multirow{3}{*}{ 0.03 }\\
 & $2$ & 0.21 & 1.2 & 10 & 0.29 & TRUE &  &  \\
 & $3$ & 0.19 & 1.3 & 12 & 0.43 & TRUE &  &  \\[1mm] 
\arrayrulecolor{lightgray}\hline
 \multicolumn{9}{l}{True lengthscales: $\ell_{1_{GP}}=0.11$, \,$\ell_{2_{GP}}=0.32$, \, $\ell_{3_{GP}}=0.43$ }
\end{tabular}
\caption{Iterative steps in HSGP model fitting on the $3$D data. The diagnostic procedure updating the parameters minimum $\ell$, $c$, $m$, estimated $\hat{\ell}$m and diagnostic $\hat{\ell} + 0.01 \geq \ell$ is applied on the three input dimensions. The full diagnostic process is applied until two consecutive \textit{trues} in the diagnostic are achieved for each dimension. RMSE and RMSE* refer to the root mean squared error of the HSGP computed against the true function and the exact GP, respectively.}
  \label{tab_caseIII_3D}
\end{table}

@@

Figure \ref{ch5_fig16_errors_exII} shows the difference  between the true underlying data-generating function and the posterior mean for GP, HSGP and spline-based model solutions, respectively. The splines-based model was fitted using a cubic spline basis, penalized by the conventional integrated square second derivative cubic spline penalty \citep{wood2017generalized}, and implemented in the R-package \textit{mgcv} \citep{wood2011mgcv}. A Bayesian approach is used to fit this spline-based model using the R-package \textit{brms} \citep{burkner2017brms}. For the spline-based model, $40$ knots in each dimension were used. Figure \ref{ch5_fig16_errors_exII}-left shows the root mean squared error (RMSE), computed against the data-generating function, as a function of the boundary factor $c$ and number of univariate basis functions $m$, for the HSGP, and knots, for the $2$D spline-based model. It can be seen that the posterior mean for HSGP and exact GP accurately represents the true data-generating function. In contrast, the accuracy of the spline-based model is significantly worse.

@@

Figure~\ref{fig_time_nD_log} shows the computation time for dynamic HMC (in seconds per iteration) in $2$D, $3$Dm and $4$D input spaces and different sized data sets, $n=300, n=500, n=1000$, and $n=3000$, as a function of the number of multivariate basis functions $m^*=m_1\times \cdots \times m_D$ used in the approximation.
Both the time of computation and $m^*$ are plotted in the logarithmic scale.

@@

For a very wiggly $2$D function (say, with $\ell_1=0.1$ and $\ell_2=0.1$), the approximate number of multivariate basis functions needed is $m*=22\times 22= 484$, which results in significantly faster computation than the regular GP, even with small data sets (i.e., $n<300$) (see Figure~\ref{fig_time_nD_log}). For a very wiggly $3$D function (say, with $\ell_1=0.1$, $\ell_2=0.1$ and $\ell_3=0.1$), the approximate number of multivariate basis functions needed is $m*=22\times 22\times 22= 10648$, which, unless for small data sets (i.e., $n<1000$), the method is still significantly faster than the regular GP. A $3$D function where each dimension has a lengthscale around $0.1$ is not that common in statistical analysis, and thus in many cases the approximation will be significantly faster than excat GP. For $4$D data sets, the method can still be more efficient than the exact GP for moderate-to-large data sets (i.e., $n>1000$).

@@

The Stan model codes for the exact GP, the HSGPm and the spline-based models of this case study can be found online at {\small \url{https://github.com/gabriuma/basis_functions_approach_to_GP/tree/master/Paper/Case-study_2D-Simulated-data}}\,.

>>>>

Equations in Section~\ref{sec_num_equations} to update $m$ and $c$ imply:
% 
\begin{itemize}
	\item $c$ must be big enough for a given $\ell$, and
	\item $m$ must be big enough for given $\ell$ and $c$. 
\end{itemize}
%
If larger than minimal $c$ and $m$ (for a given $\ell$) are used in the initial HSGP model, it is likely that the results are already sufficiently accurate. As $\ell$ is initially unknown, we recommend using this diagnostic in an iterative procedure by starting with $c$ and $m$ based on some initial guess about $\ell$, and if the estimated $\hat{\ell}$ is below the diagnostic threshold, select new $c$ and $m$ using $\hat{\ell}$. This can be repeated until
%
\begin{itemize}
\item the estimated $\hat{\ell}$ is larger than the diagnostic threshold given $c$ and $m$, and
\item the predictive accuracy measures, for example, root mean square error (RMSE), coefficient of determination ($R^2$), or expected log predicitve density (ELPD) do not improve.
\end{itemize}

As commented above, the estimated $\hat{\ell}$ being larger than diagnostic threshold does not guarantee that the approximate is sufficiently accurate, and thus we recommend to look at the predicitve accuracy measures, too.

Apart from providing a powerful diagnostic tool in determining if the approxiamtion is sufficiently accurate, the equations in the previous Section~\ref{sec_num_equations} also provide the optimal values for $m$ (the minimum $m$ required for an accurate approximation) and $c$ (the minimum $c$ that allows for the minimum $m$) that van be used to minimize the computational cost in repeated computations (e.g., in cross-validation and simulation based calibration)
%If the initial lengthscale is not far from the optimal, and larger than minimal $c$ and $m$ are used in the initial HSGP model, it is likely that the initial results are already sufficiently accurate. 
%The iterative procedure, briefly commented above and detailed in next Section~\ref{sec_user_guide}, works as safety approach and also allows finding a smaller sufficient $m$ which can be useful if the user wants to optimize the computational speed.
%Using the optimum values for $m$ and $c$ that allow for minimum computational requirements can be essential for repeated fitting (e.g., in cross-validation or simulation based calibration), uni-dimensional functions with very large data sets, and
This is even more useful in multi-dimensional cases ($2 \leq D \leq 4$), where knowing the smallest useful value of $m$ for each dimension has even bigger effect on the total computational cost.

@@

The HSGP model is computationally superior in $1$D and $2$D even for highly wiggly functions, except when the number of data points is so small ($n \lesssim 300$) that exact GPs are already reasonably fast themselves. However, the computation time of the HSGP model increases rapidly with the number of input dimensions ($D$) since the number of multivariate basis functions $m^*=m_1\times \cdots \times m_D$ in the approximation increases exponentially with $D$ (see eq.~(\ref{eq_m_multi})). Yet, the HSGP method can still be computationally faster than the exact GP for larger datasets due the latter's cubic scaling in $n$.

In our experiments of multivariate problems (see Section~\ref{sec_computation_nD}), the computation time for the HSGP model was faster than for the exact GP for most of the non-linear $2D$ functions or moderate-to-large sized $3$D datasets ($n \gtrsim 1000$), even for highly wiggly 3D functions (e.g., $\ell_1, \ell_2, \ell_3 \approx 0.1$).

{\color{blue} For small sized datasets ($n\lesssim 1000$), HSGPs are likely to be slower than exact GPs already for highly to moderated wiggly $3$D functions (e.g., $\ell_1 \lesssim 0.1$, and $\ell_2, \ell_3 \gtrsim 0.3$) and for overall smooth $4$D functions (e.g., $\ell_1 \lesssim 0.1$, and $\ell_2, \ell_3, \ell_4 \gtrsim 0.4$).}

@@

As it has been shown in case study III (Section~\ref{sec_caseIII}), the proposed diagnostic tool can be very useful for multivariate problems as it allows one to reduce $m^*$ to the minimum sufficient value, reducing computational time drastically, and still getting an accurate approximation. {\color{blue} For example, assuming a squared exponential covariance function, choosing the optimal value for $c$ allows one to use few basis functions in every single dimension (i.e., $m \lesssim 10$ for $\ell \gtrsim 0.3$; $20 \gtrsim m \gtrsim 10$ for $0.3 \gtrsim \ell \gtrsim 0.1$; and $m \gtrsim 20$ for $\ell \lesssim 0.1$), which, from results presented in Figure~\ref{fig_time_nD_log}, implies that the HSGP model can be, in general terms, useful for highly wiggly $3$D functions and smooth $4$D functions.}

@@

In our experiments, we observed roughly similar or even worse scaling behavior for the thin plate splines in comparison to HSGPs, with serious challenges with computation time when doing inference with spline models for $D=3$ and more than 10 basis functions per dimension, or even for $D=2$ and more than 40 basis functions per dimension (see Figures~B.3-right and C.3 in the online supplement).

@@

The joint posterior parameter distributions are estimated by sampling using the dynamic HMC algorithm implemented in Stan \citep{StanTeam:2021}. $\mathrm{Normal}(0,1)$, \linebreak $\mathrm{Normal}(0,3)$ and $\mathrm{Gamma}(1.2,0.2)$ prior distributions has been used for the observation noise $\sigma$, covariance function marginal variance $\alpha$, and length-scale $\ell$, respectively. We use the same prior distributions to fit the exact GP model.

@@

The Stan model code for the exact GP, the approximate GP and the spline models of this case study can be found online at {\small \url{ https://github.com/gabriuma/}} \linebreak {\small \url{ basis_functions_approach_to_GP/tree/master/Paper/Case-study_1D-Simulated-data}}\,.

@@

The joint posterior parameter distributions are estimated by sampling using the dynamic HMC algorithm implemented in Stan \citep{StanTeam:2021}. $\mathrm{Normal}(0,1)$, \linebreak $\mathrm{Normal}(0,10)$ and $\mathrm{Normal}(0,2)$ prior distributions has been used for the observation noise $\sigma$, covariance function marginal variances $\bm{\alpha}=\{\alpha_1 \alpha_2, \alpha_3\}$, and length-scales $\bm{\ell}=\{\ell_1, \ell_2, \ell_3\}$, respectively. A $\mathrm{Normal}(0,0.1)$ prior distribution has been used for the standard deviation $\tau$ of the Student's $t$ distribution with 1 degree of freedom used to model $f_4$ (i.e., the special days effects).

@@

The Stan model code for the approximate GP model of this case study can be found online at {\small \url{https://github.com/gabriuma/basis_functions_approach_to_GP/tree/}} \linebreak {\small \url{master/Paper/Case-study_Birthday-data}}\,.

@@

In this case study, we apply the diagnostic tool to fit and diagnose two different data sets, one data set simulated from a bivariate (D=$2$) function and another data set simulated from a $3$-dimensinal (D=$3$) function. Furthermore, in Section~\ref{sec_computation_nD} we present results of the computational time required to fit the HSGP model in $2$D, $3$D and $4$D input spaces, with different sized data sets and as a function of the number of multivariate basis functions $m^*=m_1\times \cdots \times m_D$ used in the approximation.

{\color{blue} $2$D and $3$D synthetic functions were drawn from $2$D and $3$D GP priors, with input values $\bm{x}_i \in [-1,1]^2$ and $\bm{x}_i \in [-1,1]^3$, respectively. Squared exponential covariance functions with marginal variance $\alpha=1$ and length-scales $\ell_1=0.10$, $\ell_2=0.3$, and $\ell_3=0.4$, where $\ell_i$ is the length scale for the $i$'th dimension, were used in the GP priors. $200$ and $1000$ data points were sampled from the $2$D and $3$D drawn functions, respectively, and independent Gaussian noise with standard deviation $\sigma=0.2$ was added to the data points to form the final noisy sets of observations.}

@@

\begin{table}
\centering
\setlength{\tabcolsep}{3.6pt}
\begin{tabular}{cccccccccc}
\arrayrulecolor{gray}\hline \\[-3mm]
 iter. & Dim. & $\ell$ & $c$ & $m$ & $\hat{\ell}$ & $\hat{\ell} + 0.01 \geq \ell$ & RMSE & RMSE* \\ 
\arrayrulecolor{lightgray}\hline \\[-2mm]
 \multirow{2}{*}{ 1 } & $1$ & 0.50 & 1.6 & 6 & 0.19 & FALSE & \multirow{2}{*}{ 0.75 } & \multirow{2}{*}{ 0.72 }\\ 
 & $2$ & 1.00 & 3.2 & 6 & 0.25 & FALSE &  &  \\
\arrayrulecolor{lightgray}\hline \\[-2mm]
 \multirow{2}{*}{ 2 } & $1$ & 0.19 & 1.2 & 12 & 0.11 & FALSE  & \multirow{2}{*}{ 0.18 } & \multirow{2}{*}{ 0.12 }\\
 & $2$ & 0.25 & 1.2 & 9 & 0.25 & TRUE &  &  \\
\arrayrulecolor{lightgray}\hline \\[-2mm]
 \multirow{2}{*}{ 3 } & $1$ & 0.11 & 1.2 & 20 & 0.11 & TRUE  & \multirow{2}{*}{ 0.11 } & \multirow{2}{*}{ 0.03 }\\
 & $2$ & 0.19 & 1.2 & 11 & 0.27 & TRUE &  &  \\
\arrayrulecolor{lightgray}\hline \\[-2mm]
 \multirow{2}{*}{ 4 } & $1$ & 0.10 & 1.2 & 22 & 0.11 & TRUE  & \multirow{2}{*}{ 0.10 } & \multirow{2}{*}{ 0.02 }\\
 & $2$ & 0.19 & 1.2 & 11 & 0.27 & TRUE  &  & \\[1mm] 
\arrayrulecolor{lightgray}\hline
 \multicolumn{8}{l}{{\small Exact GP length-scales:} $\ell_{1_{GP}}=0.10$, \, $\ell_{2_{GP}}=0.29$ }
\end{tabular}
\caption{Iterative steps in HSGP model fitting on the $2$D data. The diagnostic procedure updating the parameters minimum $\ell$, $c$, $m$, estimated $\hat{\ell}$, and diagnostic $\hat{\ell} + 0.01 \geq \ell$ is applied on the two input dimensions. The full diagnostic process is applied until two consecutive \textit{trues} in the diagnostic are achieved for each dimension. RMSE and RMSE* refer to the root mean squared error of the HSGP computed against the true function and the exact GP, respectively.}
  \label{tab_caseIII_2D}
\end{table}

\begin{table}
\centering
\setlength{\tabcolsep}{3.6pt}
\begin{tabular}{cccccccccc}
\arrayrulecolor{gray}\hline \\[-3mm]
 iter. & Dim. & $\ell$ & $c$ & $m$ & $\hat{\ell}$ & $\hat{\ell} + 0.01 \geq \ell$ & RMSE & RMSE* \\ 
\arrayrulecolor{lightgray}\hline \\[-2mm]
 \multirow{3}{*}{ 1 } & $1$ & 0.50 & 1.6 & 6 & 0.14 & FALSE & \multirow{3}{*}{ 0.63 } & \multirow{3}{*}{ 0.58 }\\ 
 & $2$ & 0.50 & 1.6 & 6 & 0.27 & FALSE &  &  \\
 & $3$ & 1.00 & 3.2 & 6 & 0.23 & FALSE &  &  \\
\arrayrulecolor{lightgray}\hline \\[-2mm]
 \multirow{3}{*}{ 2 } & $1$ & 0.13 & 1.2 & 16 & 0.11 & FALSE  & \multirow{3}{*}{ 0.13 } & \multirow{3}{*}{ 0.07 }\\
 & $2$ & 0.27 & 1.2 & 8 & 0.29 & TRUE &  &  \\
 & $3$ & 0.23 & 1.2 & 19 & 0.40 & TRUE &  &  \\
\arrayrulecolor{lightgray}\hline \\[-2mm]
 \multirow{3}{*}{ 3 } & $1$ & 0.11 & 1.2 & 19 & 0.12 & TRUE  & \multirow{3}{*}{ 0.11 } & \multirow{3}{*}{ 0.04 }\\
 & $2$ & 0.21 & 1.2 & 10 & 0.29 & TRUE &  &  \\
 & $3$ & 0.19 & 1.3 & 12 & 0.43 & TRUE &  &  \\
 \arrayrulecolor{lightgray}\hline \\[-2mm]
 \multirow{3}{*}{ 4 } & $1$ & 0.10 & 1.2 & 21 & 0.12 & TRUE  & \multirow{3}{*}{ 0.11 } & \multirow{3}{*}{ 0.03 }\\
 & $2$ & 0.21 & 1.2 & 10 & 0.29 & TRUE &  &  \\
 & $3$ & 0.19 & 1.3 & 12 & 0.43 & TRUE &  &  \\[1mm] 
\arrayrulecolor{lightgray}\hline
 \multicolumn{9}{l}{{\small Exact GP length-scales:} $\ell_{1_{GP}}=0.11$, \,$\ell_{2_{GP}}=0.32$, \, $\ell_{3_{GP}}=0.43$ }
\end{tabular}
\caption{Iterative steps in HSGP model fitting on the $3$D data. The diagnostic procedure updating the parameters minimum $\ell$, $c$, $m$, estimated $\hat{\ell}$m and diagnostic $\hat{\ell} + 0.01 \geq \ell$ is applied on the three input dimensions. The full diagnostic process is applied until two consecutive \textit{trues} in the diagnostic are achieved for each dimension. RMSE and RMSE* refer to the root mean squared error of the HSGP computed against the true function and the exact GP, respectively.}
  \label{tab_caseIII_3D}
\end{table}

@@

{\color{blue} Figure \ref{ch5_fig16_errors_exII} shows the difference  between the true underlying data-generating function and the posterior mean for GP, HSGP and spline-based model solutions for the $2$D data set. For the $3$D data set plots with the posterior functions are not shown because it is difficult to plot functions in a $3$D input space.} The splines-based model was fitted using a cubic spline basis, penalized by the conventional integrated square second derivative cubic spline penalty \citep{wood2017generalized}, and implemented in the R-package \textit{mgcv} \citep{wood2011mgcv}. A Bayesian approach is used to fit this spline-based model using the R-package \textit{brms} \citep{burkner2017brms}. For the spline-based model, $40$ knots in each dimension were used. Figure \ref{ch5_fig16_errors_exII}-left shows the root mean squared error (RMSE), computed against the data-generating function, as a function of the boundary factor $c$ and number of univariate basis functions $m$, for the HSGP, and knots, for the $2$D spline-based model. It can be seen that the posterior mean for HSGP and exact GP accurately represents the true data-generating function. In contrast, the accuracy of the spline-based model is significantly worse.

@@

Figure~\ref{fig_time_nD_log} shows the computation time for dynamic HMC (in seconds per iteration) in $2$D, $3$D and $4$D input spaces and different sized data sets, $n=300, n=500, n=1000$, and $n=3000$, as a function of the number of multivariate basis functions $m^*=m_1\times \cdots \times m_D$ used in the approximation.
Both the time of computation and $m^*$ are plotted in the logarithmic scale.

@@

For a very wiggly $2$D function (say, with $\ell_1=0.1$ and $\ell_2=0.1$), the approximate number of multivariate basis functions needed is $m^*=22\times 22= 484$, which results in significantly faster computation than the exact GP, even with small data sets (i.e., $n \lesssim 300$) (see Figure~\ref{fig_time_nD_log}). For a very wiggly $3$D function (say, with $\ell_1=0.1$, $\ell_2=0.1$ and $\ell_3=0.1$), the approximate number of multivariate basis functions needed is $m^*=22\times 22\times 22= 10648$, which, unless for small data sets (i.e., $n \lesssim 1000$), the method is still significantly faster than the regular GP. A $3$D function where each dimension has a lengthscale around $0.1$ is not that common in statistical analysis, and thus in many cases the approximation will be significantly faster than excat GP. For $4$D data sets, the method can still be more efficient than the exact GP for moderate-to-large data sets (i.e., $n \lesssim 1000$).

@@

The Stan model codes for the exact GP, the HSGP and the spline-based models of this case study can be found online at {\small \url{https://github.com/gabriuma/}} \linebreak {\small \url{basis_functions_approach_to_GP/tree/master/Paper/Case-study_2D-Simulated-data}}\,.
